---
title: "Assignment 2: Base R (8 marks)"
output:
    html_document:
        toc: false
---

```{r, echo=FALSE}
# This removes the answers for the assignments, while keeping the assignment
# contents.
knitr::opts_chunk$set(eval = FALSE, echo = FALSE)
```

*To submit this assignment, upload the full document on blackboard, including
the original questions, your code, and the output. Submit both your `.Rmd`
source file and a knitted `.html` version of the same file.*


1. Variable assignment (1 mark)

    a. Assign the value `5` to the variable/object `a`. Display `a`. (0.25
    marks)

    ```{r}
    a <- 5
    a
    ```

    b. Assign the result of `10/3` to the variable `b`. Display `b`. (0.25
    marks)

    ```{r}
    b <- 10/3
    b
    ```

    c. Assign the product of `a` and `b` to `c`. Display `c`. (0.25 marks)

    ```{r}
    c <- a * b
    c
    ```

    d. Write a function that adds two numbers and returns their sum. Use it to
    assign the sum of `a` and `b` to `c`. Display `c`. (In practice, there is
    already a more sophisticated built-in function for this: `c <- sum(a, b)`)
    (0.25 marks)

    ```{r}
    add_two_values <- function(arg1, arg2) {
        return(arg1 + arg2)
    }
    c <- add_two_values(a, b)
    c
    ```

2. Vectors (1 mark)

    a. Create a vector `v` with all integers 0-30, and a vector `w` with every
    third integer in the same range. (0.25 marks)

    ```{r}
    v <- seq(0, 30) # c(1:30)
    v

    w <- seq(0, 30, 3)
    w
    ```

    b. What is the difference in lengths of the vectors `v' and `w`? (0.25
    marks)

    ```{r}
    length(v) - length(w)
    ```

    c. Create a new vector, `v_square`, with the square of elements 3, 6, 7,
    10, 15, 22, 23, 24, and 30 from the variable `v`. _Hint: Use indexing rather
    than a for loop._ (0.25 marks)

    ```{r}
    v_square <- v[c(3,6,7,10,15,22,23,24,30)]^2
    v_square
    ```

    d. Calculate the mean and median of the first five values from `v_square`.
    (0.25 marks)

    ```{r}
    mean(v_square[1:5])
    median(v_square[1:5])
    ```

3. Boolean indexing (1 mark)

    a. Create a boolean vector `v_bool`, indicating which vector elements are
    bigger than 20. How many values are over 20? _Hint: In R, TRUE = 1, and
    FALSE = 0, so you can use simple arithmetic to find this out._ (0.5 marks)

    ```{r}
    v_bool <- v > 20
    sum(v_bool)
    ```

    b. Use the variable `v_bool` as an index to extract the elements from `v`
    that are bigger than 20. What are the min and max values of this new
    vector? (0.5 marks)

    ```{r}
    min(v[v_bool])
    max(v[v_bool])
    ```

4. Data frames (2 marks)

    a. There are many built-in data frames in R, which you can find [more
    details about
    online](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/00Index.html).
    What are the column names of the built-in dataframe `beaver1`? How many
    observations (rows) and variables (columns) are there? (0.5 marks)

    ```{r}
    colnames(beaver1)
    dim(beaver1) # nrow(iris)
    ```

    b. How can you view the first 5 rows of this data frame? How can you view
    the last 5 rows? (0.5 marks)

    ```{r}
    head(beaver1)
    tail(beaver1)
    ```

    c. What is the min, mean, and max body temperatue in this data set? _Hint:
    Remember that each column in a data frame is a vector, so you can use the
    same functions as in the previous question on vectors._ (0.5 marks)

    ```{r}
    min(beaver1$temp)
    mean(beaver1$temp)
    max(beaver1$temp)
    # summary(beaver1$temp)
    ```

    d. Use the `summary` function to display an overview of the `temp` column.
    (0.5 marks)

    ```{r}
    summary(beaver1$temp)
    ```

5. Data frames with dplyr (3 marks)

    a. Use dplyr to calculate the mean temperature in the dataset. (0.25 marks)

    ```{r}
    library(dplyr)
    beaver1 %>%
        summarize(mean_temp = mean(temp))
    ```

    b. Calculate the mean temperature for day 346. (0.5 marks)

    ```{r}
    beaver1 %>%
        filter(day == 346) %>%
        summarize(mean_temp_346 = mean(temp))
    ```

    c. Rather than using `filter()` to calculate the mean for each day
    separately, the more convenient `group_by()` can be used to aggregate
    measurements by a categorical value (such as the `day` column in `beaver`).
    Use this approach to calculate the mean temperature and activity level for
    each of the days in the dataset. (0.5 marks)

    ```{r}
    beaver1 %>%
        group_by(day) %>%
        summarize(mean_temp = mean(temp), mean_activity = mean(activ))
    ```

    d. Express in writing what the average activity level from the above
    calculation means. _Hint: Remember that you can [read a description of the
    columns
    online](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/00Index.html)._
    (0.25 marks)

    ```{r}
    # Expected answer:
    # The beaver was active outside the retreat for 5.5 % of the
    # observations on day 346 and 4.3 % of the observations on day 347.
    ```

    e. How many observations are there per day in this dataset? (0.25 marks)

    ```{r}
    beaver1 %>%
        count(day)
    ```

    f. How many observations are there per day when the beaver is active
    outside the retreat? (0.25 marks)

    ```{r}
    beaver1 %>%
        filter(activ == 1) %>%
        count(day)
    ```

    g. Was the body temperature higher when the beaver was active or inactive?
    Is this what you would expect? (0.25 marks)

    ```{r}
    beaver1 %>%
        group_by(activ) %>%
        summarize(mean_temp = mean(temp))
    ```

    h. Add to your solution in `f` by also grouping by activity level *and* the
    day of the observation. Which variable seems to be more correlated with
    high body temperature, activity level or the day of the measurment? (0.25
    marks)

    ```{r}
    beaver1 %>%
        group_by(activ, day) %>%
        summarize(mean_temp = mean(temp))
    ```
