---
title: 'Assignment: Data visualization (8 marks)'
output:
    html_document:
        toc: false
---

<!-- keep this part -->
*To submit this assignment, upload the full document on blackboard,
including the original questions, your code, and the output. Submit both
your `.Rmd` source file and a knitted `.html` version of the same file.*

1.  Recreate a customized plot using what you've learned (3.5 marks).

    We will use the CO2 dataset studying the cold tolerance on the grass
    species *Echinochloa crus-galli*.

```{r}
# Load the CO2 dataset in base R (there is also a co2 dataset, be careful with capitalization)
CO2 <- datasets::CO2
```

    Recreate the plot below using the following specified parameters:

    - colors: grey & orange (*Hint:* Use `scale_color_manual`. Check the help docs)
    - point size: 3
    - point alpha: 0.6
    - theme: theme_light
    - smoothing method: 'loess' (default)
    - smoothing function alpha: 0.2
    
```{r, echo=FALSE, message=FALSE}
library(tidyverse)

CO2 <- datasets::CO2
color_names <- c('gray', 'orange')
ggplot(CO2, aes(x = conc, y = uptake, color = Treatment)) + 
    geom_point(alpha = 0.6) + 
    geom_smooth(method = 'loess', alpha = 0.2) + 
    facet_wrap( ~ Type) +
    scale_color_manual(values = color_names) + 
    theme_light() +
    theme(legend.position = 'top') +
    xlab("Concentration") + 
    ylab("CO2 Uptake")
```


2.  Create your own custom ggplot2 theme for your team. You'll use this
    theme when making all of your graphs for your final report. Work
    with your team, using the code below as the base, to decide how you
    want all your graphs to look like. What we exactly expect as an
    answer isn't specific, but we do expect you and your team to try to
    make an effort at making your own custom and visually appealling
    theme. See the links below for help and documentation on changing
    all aspects of your graph theme. Paste the code in this question. (4
    marks)

```{r}
library(ggplot2)
# Sample data and plot to use to create your theme
p <- ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = cut(Petal.Length, 3))) +
    geom_point() +
    facet_grid( ~ Species)

team_theme <- function() {
    # Make all your edits here for the theme.
    # The code below is an example, change them when creating your theme
    theme(axis.line = element_line(color = "red"),
          text = element_text(size = 8),
          legend.text = element_text(size = 12, family = "serif")) # this is an example
}
p + team_theme()
```

-   ggplot2 official documentation on all the possible theme items:
    <http://ggplot2.tidyverse.org/reference/theme.html>
-   ggplot2 official documentation on modifying theme items:
    <http://ggplot2.tidyverse.org/reference/theme_get.html>
-   Great online book with tutorial walking through how to create a new
    theme:
    <https://bookdown.org/rdpeng/RProgDA/building-a-new-theme.html#creating-a-new-theme>
-   ggthemes (package with even more pre-created themes) documentation
    on what possible themes look like (can use this as inspiration, and
    even copy if you want! But don't copy the themes entirely... you
    want the theme to be your own):
    <https://cran.r-project.org/web/packages/ggthemes/vignettes/ggthemes.html>

3.  Once you've created the custom theme, decide on a team member that
    will save this theme to their fork of your team's GitHub account.
    Create a new file in the `R/` folder called `graph_theme.R`. Paste
    the `team_theme` function code into this file. Commit this file and
    submit a PR of this change to the main team repo. (0.5 marks)
    
    - URL of pull request:
