---
title: 'Assignment 5: Linear mixed-effects models and randomization tests (8 marks)'
output:
    html_document:
        toc: false
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

*To submit this assignment, upload the full document on blackboard,
including the original questions, your code, and the output. Submit
your assignment as a knitted `.pdf` (preferred) or `.html` file.*

1. Linear mixed-effects models (4 marks).

    Santangelo _et al._ (In press) were interested in understanding how plant
    defenses, herbivores, and pollinators influence the expression of plant floral
    traits (e.g. flower size). Their experiment had 3 treatments, each with 2
    levels: Plant defense (2 levels: defended vs. undefended), herbivory (2 levels:
    reduced vs. ambient) and pollination (2 levels: open vs. supplemental). These
    treatments were fully crossed for a total of 8 treatment combinations. In each
    treatment combination, they grew 4 individuals from each of 25 plant genotypes
    for a total of 800 plants (8 treatment combinations x 25 genotypes x 4
    individuals per genotype). Plants were grown in a common garden at the Koffler
    Scientific Reserve (UofTs field research station) and 6 floral traits were
    measured on all plants throughout the summer. We will analyze how the treatments
    influenced one of these traits in this exercise. Run the code chunk below to
    download the data, which includes only a subset of the columns from the full
    dataset:
    
    ```{r}
    library(tidyverse)
    
    plant_data <- "https://uoftcoders.github.io/rcourse/data/Santangelo_JEB_2018.csv"
    download.file(plant_data, "Santangelo_JEB_2018.csv")
    plant_data <- read_csv("Santangelo_JEB_2018.csv", 
                           col_names = TRUE)
    glimpse(plant_data)
    head(plant_data)
    ```

    You can see that the data contain 792 observations (i.e. plants, 8 died during
    the experiment). There are 50 genotypes across 3 treatments: Herbivory,
    Pollination, and HCN (i.e. hydrogen cyanide, a plant defense). There are 6 plant
    floral traits: Number of days to first flower, banner petal length, banner petal
    width, plant biomass, number of flowers, and number of inflorescences. Finally,
    since plants that are closer in space in the common garden may have similar
    trait expression due to more similar environments, the authors included 6
    spatial "blocks" to account for this environmental variation (i.e. Plant from
    block A "share" an environment and those from block B "share" an environment,
    etc.). Also keep in mind that each treatment combination contains 4 individuals
    of each genotype, which are likely to have similar trait expression due simply
    to shared genetics.
    
    a. Use the `lme4` and `lmerTest` R packages to run a linear mixed-effects model
    examining how herbivores (`Herbivory`), Pollinators (`Pollination`), plant
    defenses (`HCN`) _and all interactions_ influences the length of banner petals
    (`Avg.Bnr.Ht`) produced by plants while accounting for variation due to spatial
    block and plant genotype. Also allow the intercept for `Genotype` to vary across
    the levels of the herbivory treatment. (1 mark: 0.5 for correct fixed effects
    specification and 0.5 for correct random effects structure). You only need to
    specify the model for this part of the question.

    b. Summarize (i.e. get the output) the model that you ran in part (a). Did
    any of the treatments have a significant effect on banner petal length? If
    so, which ones? Based on your examination of the model output, how can you
    tell which level of the significant treatments resulted in longer or shorter
    mean banner petal lengths? Make a statement for each significant effect in
    the model (1 mark).

    c. Using `dplyr` and `gglot2`, plot the mean banner length for one of the
    significant interactions in the model above (whichever you choose). The idea
    is to show how both treatments interact to influence the mean length of
    banner petals (**Hint** use different shapes, colours, linetypes, etc. in
    `ggplot()`). Feel free to use whatever kind of plot you would that is
    appropriate to this kind of data. Also include 1 standard error around the
    mean. **Hint:** you should summarize the data in `dplyr` prior to plotting
    with `ggplot()`. As a reminder, I have included the formula to calculate
    the standard error of the mean below. (1 mark).

    $$ SE = \frac{sd}{\sqrt{n}}  $$

    d. After accounting for the fixed effects, how much of the variation in
    banner petal length was explained by each of the random effects in the
    model? Show your work (0.5 marks).

    e. Descibe the pattern you see in the figure generated in part (c). Why do
    you think the interaction you plotted was significant in the model? Suggest
    one plausible ecological explanation for the observed pattern. (0.5 marks)

2. Simulating data (2 marks)

    a. Generate a gamma distribution by randomly sampling 30 points from a distribution with shape parameter equal to 1.25 and rate parameter equal to 0.5. Plot this distribution. Set a seed of 42. (0.5 mark).

    b. Plot the distribution of sample means obtained by generating 5000 gamma distributions with the same parameters as in (a). In other words, the distribution should be made up of 5000 means, each from a different simulated gamma distribution. Set a seed of 43. What do you notice about this distribution when compared to the original distribution in (a)? Why would we expect this? (0.5 marks)

    c. In this exercise you will simulate a multiple regression. Remember, multiple regression means that there is more than one explanatory (aka predictor, independent) variable for a given response variable. Multiple regression thus estimates a separate effect (i.e. _beta_) for each explanatory variable in the model, while holding the other variables constant. This exercise is only a slight extension of the model that we simulated in lecture. Simulate a model that satisfies the conditions below and show the model output using `summary()`. Set a seed of 44. (1 mark).

        1. `x1` is an explanatory variable with _sequence_ from 41 to 60 with 1 unit intervals between each value (i.e. 20 values total).
        2. `x2` is an explanatory variable of length 20 randomly drawn for a normal distribution with mean equal to 62 and standard deviation equal to 2.7.
        2. `x3` is an explanatory variable of length 20 randomly drawn for a gamma distribution with shape equal to 5 and rate equal to 0.5.
        3. the `y_intercept` is 20
        4. The beta associated with `x1` is 0.62.
        5. The beta associated with `x2` is 0.047`
        6. The beta associated with `x3` is 0.175
        6. The error is drawn from a normal distribution with mean equal to 0 and standard deviation equal to 1.65.
        7. `y` is a linear combination of `x1`, `x2` and `x3`. There are no interations. Don't forget about the error. 

3. Randomization test (2 marks)

    Run the code chunk below to load the data that will be used in this exercise.

    ```{r message=FALSE, warning=FALSE}
    df <- "https://uoftcoders.github.io/rcourse/data/Assign05_Question3.csv"
    download.file(df, "Assign05_Question3.csv")
    df <- read_csv("Assign05_Question3.csv", 
                           col_names = TRUE)
    glimpse(df)
    ```

    a. Generate a histogram showing the null distribution of t-statistics between groups one and two from the `df` dataframe that you just loaded. The null distribution should be based 5,000 reshufflings of the data. (1 mark). Overlay onto this histogram the observed t-statistic as a dashed vertical line. Set a seed of 46. **Hint:** You'll need to extract the t stat from a t-test saved as an object.

    b. Perform a permutation test testing whether the observed t statistic between groups one and two is different than what would be expected by chance. Include a statement about whether there is a significant difference between groups based on your permutation test and be sure to include the P-value. How does this P-value compare to one obtained from a simple t-test? Why? (1 mark)
