---
title: "Introduction to course"
author: "Joel Östblom & Ahmed Hasan"
---

```{r echo=FALSE, results='hide', message=FALSE, warning=FALSE}
# Load this before class starts
library(tidyverse)
library(GGally)
```

## Lesson preamble

> ### Lesson objectives
> 
> - **Main goal**:
>     - To give students an overview of the capabilities of R and how to use the
>       R Notebook for exploratory data analyses.
> - **Secondary goals**:
>     - Learn about the differences between R and Excel.
>     - Learn some basic R commands.
>     - Learn about the markdown syntax and how to use it within the R Notebook.
> 
> ### Lesson outline
> 
> Total lesson time: 1 hour
> 
> - Working with computers (10 min)
> - Programming in R (45 min)
>     - The Basics (10 min)
>     - RStudio and the R Notebook (5 min)
>     - Creating and assigning objects (5 min)
>     - Plotting (10 min)
>     - Saving data and generating reports (5 min)
>     - R Markdown (10 min)

-----

## Working with computers

Before we get into more practical matters, I want to provide a brief background
to the idea of working with computers. Essentially, computer work is about
humans communicating with a computer by modulating flows of current in the
hardware in order to get the computer to carry out advanced calculations that we are
unable to efficiently compute ourselves. Early examples of human computer
communication were quite primitive and included physically disconnecting a wire
and connecting it again in a different spot. Luckily, we are not doing this
anymore; instead we have programs with graphical user interfaces with menus and
buttons that enable more efficient human to computer communication.

### Graphical user interfaces vs text based user interfaces

An example of such a program that I think many of you are familiar with is
spreadsheet software such as Microsoft Excel and LibreOffice Calc. Here, all
the functionality of the program is accessible via hierarchical menus, and
clicking buttons sends instructions to the computer, which then responds and sends
the results back to your screen. For instance, I can click a button to send the
instruction of coloring this cell yellow, and the computer interprets my
instructions and then displays the results on the screen; in this case, the
cell is highlighted yellow.

Spreadsheet software is great for viewing and entering small data sets and
creating simple visualizations fast. However, it can be tricky to design
publication-ready figures, create automatic reproducible analysis workflows,
perform advanced calculations, and reliably clean data sets. Even when using
a spreadsheet program to record data, it is often beneficial to have some some
basic programming skills to facilitate the analyses of those data.

Typing commands directly instead of searching for them in menus is a more
efficient approach to communicate with the computer and a powerful way of doing
data analysis. This is initially intimidating for almost all people, but if you
compare it to learning a new language, the process becomes more intuitive: when
learning a language, you would initially string together sentences by looking
up individual words in the dictionary. As you improve, you will only reference
the dictionary occasionally since you already know most of the words.
Eventually, you will throw the dictionary out altogether because it is faster and more
precise to speak directly. In contrast, graphical programs force you to look up
every word in the dictionary every time, even if you already know what to say.

### Formal and natural languages

You might think it is quite tricky to learn computer languages, but it is
actually not! You already know one: mathematics is often written the same way
as you would write it by hand, both in Excel and R.

```{r}
d = 4 + 5
```

This is much more efficient than typing "Hi computer, could you please add
4 and 5 for me?". 
Formal computer languages also avoid the ambiguity present
in natural languages such as English. You can think of R as a combination of
math and a formal, succinct version of English.

E.g. Create a sequence of the numbers zero to thirty.

```{r}
seq(0, 30)
```

E.g. Create a sequence of the numbers zero to thirty and calculate the average
value of this sequence.

```{r}
mean(seq(0, 30))
```

In my experience, learning programming really is similar to learning a foreign
language - you will often learn the most from just trying to do something and
receiving feedback (from the computer or another person)! When there is
something you can't wrap you head around, or if you are actively trying to find
a new way of expressing a thought, then look it up, just as you would with
a natural language.

## Programming in R

### The basics

With that background to some of the concepts of programming, let's compare the
workflow of analyzing a data set in R vs in Excel. _Pull up Excel and an
R console next to each other._ In Excel, we open a file via 'File -> Open ->
Navigate to file'. You can conceptualize this as if there is a function
`file.open()` for which you need to specify a file location,
(`file.open(/home/joel/iris.csv`). In R this looks similar, it is
`read.csv('/home/joel/iris.csv')`, <!--need to write in windows paths...--> so
it is pretty much the same. The only differences are that slightly different words are used,
and you type them in instead of clicking your way to them.

When you open a file in Excel, it will immediately display the content in the
window. Likewise, the R console will display the information of the data set
when you read it in. This sample data set that we are using for this class
describes the length and width of sepals and petals for three species of iris
flowers.

```{r, eval=FALSE}
read_csv('iris.csv')
```

However, to do useful and interesting things to data, we need to assign _values_ to
_objects_. Objects are also referred to as _variables_, and although there
technically are minor differences between them[^objects_variables], we will use
them interchangeably in this class. To create an object, we need to give it a
name followed by the assignment operator `<-`, and the value we want to assign:

[^objects_variables]: https://cran.r-project.org/doc/manuals/r-release/R-lang.html#Objects).

```{r}
x <- 3
x + 5
```

`<-` is the assignment operator. It assigns values on the right to objects on
the left. So, after executing `x <- 3`, the value of `x` is `3`. The arrow can
be read as 3 **goes into** `x`.

Let's save our data into an object called `iris`

```{r}
iris <- read_csv('data/iris.csv')
```

```{r}
iris
```

These values are now stored in this R session and we can access them by the name `iris`.
More formally, we can say 'the data has been assigned to the variable `iris`'.

The arrow `<-` is known as the _assignment operator_ and simply means that the 
output from the right hand side is stored in the variable name that you write on the left hand side.[^equal_vs_arrow] 
Now we can simply ask R to calculate the mean of the data frame `iris`, column `petal_length`:

[^equal_vs_arrow]: For historical reasons, you can also use `=` for assignments,
but not in every context. Because of the
[slight](http://blog.revolutionanalytics.com/2008/12/use-equals-or-arrow-for-assignment.html)
[differences](https://web.archive.org/web/20130610005305/https://stat.ethz.ch/pipermail/r-help/2009-March/191462.html)
in syntax, it is good practice to always use `<-` for assignments.
Note that in RStudio, typing <kbd>Alt</kbd> + <kbd>-</kbd> will write ` <- ` in a single keystroke.

```{r}
mean(iris$petal_length)
```

The `$` is the syntax used in R to indicate that you want to access a column in
the data frame. If we want to know more than just the mean, one useful thing
with R is that we don't have to calculate these things one by one. Rather, we can
actually get an overview of all the columns by asking R to provide a summary of
`iris`.

```{r}
summary(iris)
```

You can already see how it will save you time to use R instead of typing these
calculations separately in Excel.
 
### RStudio and the R Notebook

Before we continue, we are going to switch from using the R console to using
an interface called RStudio. RStudio includes the R console, but also many
other convenient functionalities, which makes it easier to get started and to
work with R. When you launch RStudio, you will see four panels:

- The console is the same as what we have been using so far. The advantage with
  using the console within RStudio is that it interacts with the other panels.
- So, if I read in the iris data we were working with previously, we can view it
  in the console by typing `iris`, just like before, or we can also view the
  `iris` variable in our environment tab, which shows it to us in a spreadsheet
  or table format.
- This panel shows the files in the current directory, any plots we might make
  later, and also the documentation, which we accessed earlier by typing `?` in
  front of a function. Here, the documentation is formatted in a way that is
  easier to read and also provides links to the related sections.
- Finally we have the text editor panel. This is where we can write scripts,
  i.e. putting several commands of code together and saving them as a text
  document so that they are accessible for later and so that we can execute them
  all at once by running the script instead of typing them in one by one.
- Another very useful thing with RStudio is that you have access to some
  excellent cheat sheets in PDF format straight from the menu: `Help -> Cheatsheets`!

In the RStudio interface, we will be writing code in a format called the
R Notebook. As the name entails, this interface works like a notebook for code,
as it allows us to save notes about what the code is doing, the code itself,
and any output we get, such as plots and tables, all together in the same
document.

When we are in the Notebook, the text we write is normal plain text, just as if we
would be writing it in a text document. If we want to execute some R code, we
need to insert a _code chunk_. You insert a code
chunk by either clicking the "Insert" button or pressing <kbd>Ctrl</kbd> +
<kbd>Alt</kbd> + <kbd>i</kbd> simultaneously. You could also type out the
surrounding backticks, but this would take longer. To run a code chunk, you
press the green arrow, or <kbd>Ctrl</kbd> + <kbd>Shift</kbd> + <kbd>Enter</kbd>.


```{r}
seq(0:30)
```

As you can see, the output appears right under the code block. This is a great
way to perform explore your data, since you can do your analysis and write
comments and conclusions right under it all in the same document. A powerful feature
of this workflow is that there is no extra time needed for code documentation
and note-taking, since you're doing your analyses and taking notes at the same time.
This makes it great for both taking notes at lectures and to have as a reference when
you return to your code in the future. We will delve deeper into some of the
neat features about the R notebook towards the end; for now, let's get back into
learning about R.

### Plots

In R, it is straightforward to explore your data by creating nice looking
plots. For instance, to make a scatter plot, we first tell R that we want to make a plot
of the `iris` data frame, and we want the aesthetics of this plot to be that
`sepal_length` is on the x axis and `sepal_width` is on the y-axis. Finally, we
specify that we want this to be a plot of points, i.e. a scatter plot.

```{r}
ggplot(iris, aes(x = sepal_length, y = sepal_width)) +
    geom_point()
```

This plot shows up in the plot panel in RStudio, which is far more convenient than
having it show up as a separate window. You can already see that there seems to
be at least some structure to the data. There is one cluster of points up the
left corner and at least one in the lower right corner. Hmmmm, I wonder if
these correspond to the different species of flowers... Let's find out, by
coloring the points according to which species they are.

```{r}
ggplot(iris, aes(x = sepal_length, y = sepal_width, color = species)) +
    geom_point()
```

Likewise, it would be similarly easy to make a histogram of the `petal_length` and
color it according to the species.

```{r}
ggplot(iris, aes(x = petal_length, fill = species)) +
    geom_histogram()
```

And we could get fancier by instructing R to split this plot in vertically
stacked facets, or subplots, with one species in each facet.

```{r}
ggplot(iris, aes(x = petal_length, fill = species)) +
    geom_histogram() +
    facet_wrap(~ species, dir = 'v')
```

Finally, if you don't have a clear idea of what you are looking for in
a data set -- such as if you are exploring publicly available data that you have not
generated yourself -- then it can be useful to visualize the relationships
between all the variables to get an overview of the data.

```{r}
ggpairs(iris, aes(color = species))
```

This plot is a bit overwhelming, and the purpose here is not to produce
a publication ready figure. You can see that many of the labels are overlapping
and not everything fits in the plotting window. We can tweak these properties,
which we will get into later in the course. But even in its slightly rough
form, this is a powerful data set overview for your own analyses.

Hopefully, these examples have giving you an idea of how simple and powerful it
is to use R for your exploratory data analyses. In a few lines of code, we have
made some really nice plots, some of which are not even imaginable to create in
Excel.

### Saving data and generating reports

To save our code and graphs, all we have to do is to save the R Notebook file,
and the we can open it in RStudio next time again. However, if we want someone
else to look at this, we can't just send them the R Notebook file, because they might
not have RStudio installed. Another great feature of R Notebooks is that it is
really easy to export them to HTML or PDF documents with figures and
professional typesetting. There are actually many academic papers that are
written entirely in this format and it is great for assignments and reports:
I have used this myself for some of my progress updates to my supervisor. Since
R Notebook files convert to HTML, it is also easy to publish simple good-looking
websites in it -- for example, the syllabus website for this course is made this
way via a template that Luke set up.

Let's try to convert this new document. First, let's go through it and see what
we expect the output to include. First we see this intro section here -- this is called
the yaml-block, and it's where you specify the title of your document, what kind of output you want,
and a few other things that we will try later. Then we have a heading, a link, some bold text,
and a few code blocks. OK, let's see what this looks like. To create the output
document, we poetically say that we will knit our R Markdown into the HTML
document. Luckily, it is much simpler than actually knitting something. Simply
press the `Knit` button here and the new document will be created.

As you can see in the knitted document, the title showed up as we would expect,
the lines with pound sign(s) in front of them were converted into headers and
we can see both the code and its output! So the plots are generated directly in
the report without us having to cut and paste images! If we change something in
the code, we don't have to find the new images and paste it in again, the
correct one will appear right in your code.

### R Markdown

The text format we are using in the R Notebook is called R Markdown. This format allows us to combine R code with the Markdown text format, which enables the use of certain
characters to specify headings, bullet points, quotations and even 
citations. A simple example of how to write in Markdown is to use a single
asterisk or underscore to *emphasize* text (`*emphasis*`) and two asterisks or
underscores to **strongly emphasize** text (`**strong emphasis**`). When we
convert our R Markdown text to HTML or PDF, these will show up as italics and
bold typeface, respectively. If you have been writing on online forums, or
popular chat services, such as whatsapp, you might already be familiar with
this style of writing. In case you haven't seen it before, you have just learnt
something about whatsapp in your theoretical ecology class...

To get more familiar with the Markdown syntax, let's open a new R Notebook.
RStudio already includes a template when you create new documents that gives
you some instructions on how to use the syntax. We can see some of the
formatting rules that we just mentioned. Let's add some emphasized text and see
how it renders. Using the Markdown syntax we can organize our document with
headings and subheadings, each indicated by a variable number of hash symbols.

Furthermore, if we want to change the style of our output document, it is quite
easy to do so. Let's add a subheading first, and a table of contents.

```yaml
---
title: "HTML output demo"
output:
  html_document:
    toc: true
---
```

Now we realize that we want to include numbers in the report to keep with the departmental guidelines, so let's add those.

```yaml
---
title: "HTML output demo"
output:
  html_document:
    toc: true
    number_section: true
---
```

We can alter the style of our report by adding a theme.

```yaml
---
title: "HTML output demo"
output:
  html_document:
    toc: true
    number_section: true
    theme: united
---
```

And finally we remember that we need to hand in our assignments as PDF-files, so let's create a PDF instead of a HTML-document.

```yaml
---
title: "PDF output demo"
output:
  pdf_document:
    toc: true
    number_section: true
---
```

If you want to learn more about R Markdown, you can read the cheat sheets in
RStudio and [RStudio Markdown reference
online](http://rmarkdown.rstudio.com/authoring_basics.html).

## Background experience survey
Hopefully, this has given you an idea of how powerful and time-saving R can be
for your data analysis, and how we will work with it in this class using the
R Notebook. The first assignment will be available on blackboard today, and it
is includes a few questions about RStudio and R Markdown. 1 / 4 marks on this
assignment is to fill in a survey about your background knowledge. This is so
that we can adjust the pace and content of the lectures and to divide you into
groups accordingly for the class project later this fall. Please take 10 min to
do so now by going to this page (send page to everyone's computer via the
teaching software), or if you have to run, you can complete it at home.

