---
title: "Tidying and exporting data, reproducing figures"
---

> ### Lecture objectives
>
> - Learn about tidy data.
> - Transform data from the long to wide format.
> - Reproduce existing figures from raw data.
> - Understand which raw data is underlying a figure.
> - Understand which types of figures are suitable to create from raw data.
> - Explore scientific questions using dplyr and ggplot.
>
>
> ### Lecture outline
>
> - Reshaping with gather and spread (25 min)
> - Exporting data (15 min)
> - Reproducing figures (50 min)


## Setting up

Start by loading the required packages. Both **`ggplot2`** and **`dplyr`** are
included in the **`tidyverse`** package collection.

```{r}
# Install if needed
# install.packages('tidyverse')
library(tidyverse)
```

Load the data we saved in the previous lesson.

```{r}
# Download if needed
# download.file("https://ndownloader.figshare.com/files/2292169", "data/portal_data.csv")
surveys <- read_csv('portal_data.csv')
surveys
```

## Reshaping with gather and spread

**`dplyr`** is one part of a larger **`tidyverse`** that enables you to work
with data in tidy data formats. **`tidyr`** enables a wide range of
manipulations of the structure data itself. For example, the survey data
presented here is almost in what we call a **long** format - every observation
of every individual is its own row. This is an ideal format for data with a rich
set of information per observation. It makes it difficult, however, to look at
the relationships between measurements across plots. For example, what is the
relationship between mean weights of different genera across the entire data
set?

To answer that question, we'd want each plot to have a single row, with all of
the measurements in a single plot having their own column. This is called a
**wide** data format. For the `surveys` data as we have it right now, this is
going to be one heck of a wide data frame! However, if we were to summarize data
within plots and species, we might begin to have some relationships we'd want to
examine.

Let's see this in action. First, using **`dplyr`**, let's create a data frame
with the mean body weight of each genus by plot.

```{r}
surveys_gw <- surveys %>%
    filter(!is.na(weight)) %>%
    group_by(genus, plot_id) %>%
    summarize(mean_weight = mean(weight))

head(surveys_gw)
```

### Long to Wide with `spread`

Now, to make this long data wide, we use `spread` from `tidyr` to spread out the
different taxa into columns. `spread` takes three arguments: - the data, the
*key* column (or column with identifying information), the *values* column (the
one with the numbers/values). We'll use a pipe so we can ignore the data
argument.

```{r}
surveys_gw_wide <- surveys_gw %>%
  spread(genus, mean_weight)

head(surveys_gw_wide)
```

Notice that some genera have `NA` values. That's because some of those genera
don't have any record in that plot. Sometimes it is fine to leave those as
`NA`. Sometimes we want to fill them as zeros, in which case we would add the
argument `fill=0`.

```{r}
surveys_gw %>%
  spread(genus, mean_weight, fill = 0) %>%
  head
```

We can now do things like plot the weight of *Baiomys* against *Chaetodipus* or
examine their correlation.

```{r}
surveys_gw %>%
  spread(genus, mean_weight, fill = 0) %>%
  cor(use = "pairwise.complete")
```

### Wide to long with `gather`

What if we had the opposite problem, and wanted to go from a wide to long
format? For that, we use `gather` to sweep up a set of columns into one
key-value pair. We give it the arguments of a new key and value column name, and
then we specify which columns we either want or do not want gathered up. So, to
go backwards from `surveys_gw_wide`, and exclude `plot_id` from the gathering,
we would do the following:

```{r}
surveys_gw_long <- surveys_gw_wide %>%
  gather(genus, mean_weight, -plot_id)

head(surveys_gw_long)
```

Note that now the `NA` genera are included in the long format. Going from wide
to long to wide can be a useful way to balance out a dataset so every replicate
has the same composition.

We could also have used a specification for what columns to include. This can be
useful if you have a large number of identifying columns, and it's easier to
specify what to gather than what to leave alone. And if the columns are
sequential, we don't even need to list them all out - just use the `:` operator!

```{r}
surveys_gw_wide %>%
  gather(genus, mean_weight, Baiomys:Spermophilus) %>%
  head()
```

># ### Challenge
>
> 1. Make a wide data frame with `year` as columns, `plot_id` as rows, and the
>   values are the number of genera per plot. You will need to summarize before
>   reshaping, and use the function `n_distinct` to get the number of unique
>   types of a genus. It's a powerful function! See `?n_distinct` for more.
>
> 2. Now take that data frame, and make it long again, so each row is a unique
>    `plot_id` - `year` combination.
>
> 3. The `surveys` data set is not truly wide or long because there are
>    two columns of measurement - `hindfoot_length` and `weight`.  This makes it
>    difficult to do things like look at the relationship between mean values of
>    each measurement per year in different plot types. Let's walk through a
>    common solution for this type of problem. First, use `gather` to create a
>    truly long dataset where we have a key column called `measurement` and a
>    `value` column that takes on the value of either `hindfoot_length` or
>    `weight`. Hint: You'll need to specify which columns are being gathered.
>
> 4. With this new truly long data set, calculate the average of each
>    `measurement` in each `year` for each different `plot_type`. Then
>    `spread` them into a wide data set with a column for `hindfoot_length` and
>    `weight`. Hint: Remember, you only need to specify the key and value
>    columns for `spread`.

<!---
  ```{r, echo=FALSE}
## Answer 1
rich_time <- surveys %>%
  group_by(plot_id, year) %>%
  summarize(n_genera = n_distinct(genus)) %>%
  spread(year, n_genera)

head(rich_time)

## Answer 2
rich_time %>%
  gather(year, n_genera, -plot_id)

## Answer 3
surveys_long <- surveys %>%
  gather(measurement, value, hindfoot_length, weight)

## Answer 4
surveys_long %>%
  group_by(year, measurement, plot_type) %>%
  summarize(mean_value = mean(value, na.rm=TRUE)) %>%
  spread(measurement, mean_value)
```
--->

## Exporting data

Now that you have learned how to use **`dplyr`** to extract information from
or summarize your raw data, you may want to export these new datasets to share
them with your collaborators or for archival.

Similar to the `read_csv()` function used for reading CSV files into R, there is
a `write_csv()` function that generates CSV files from data frames.

Before using `write_csv()`, we are going to create a new folder,
`data-processed`, in our working directory that will store this generated
dataset. We don't want to store manipulated datasets in the same directory as
our raw data. It's good practice to keep them separate. The raw data would
ideally be put in a `data-raw` folder, which should only contain the raw,
unaltered data, and should be left alone to make sure we don't delete or modify
it from how it was when we downloaded or recorded it ourself. In contrast,
our R code will create the contents of the `data-processed` directory, so even
if the files it contains are deleted, we can always re-generate them.

Use the `getwd()` function to find out which is the current working directory.

```{r}
getwd()
```

Navigate to this directory in your file browser and create a folder called
`data-processed`.

Alternatively, you could use R to create this directory.

```{r}
dir.create("data-processed")

# To suppress the warning, we could do
dir.create("data-processed", showWarnings = FALSE)

# Another alternative would be to use a conditional expression, which only
# creates the directory *if* it does not already exist. The syntax here is
# similar to the for loop we created in the second lecture.
if (!dir.exists('data-processed')) {
    dir.create("data-processed")
}
```

We are going to prepare a cleaned up version of the data without NAs. Let's
start by removing observations for which the `species_id` is missing. Let's also
remove observations for which `weight` and the `hindfoot_length` are missing.
This dataset should also only contain observations of animals for which the sex
has been determined:

```{r}
surveys_complete <- surveys %>%
  filter(!is.na(species_id),       # remove missing species_id
         !is.na(weight),           # remove missing weight
         !is.na(hindfoot_length),  # remove missing hindfoot_length
         !is.na(sex))              # remove missing sex

# This expression is a succinct alternative to the above
surveys_complete_comcas <- surveys %>%
    filter(complete.cases(species_id, weight, hindfoot_length, sex))

# This is even briefer, but omits observations with NA in *any* column.
# There is no way to control which columns to use, but it is common to want
# to exclude all NAs, which in our case corresponds to the columns listed above.
surveys_complete_naomit <- na.omit(surveys)

# Compare the dimensions of the original and the cleaned data frame
dim(surveys)
dim(surveys_complete)
dim(surveys_complete_comcas)
dim(surveys_complete_naomit)
```

Now that our dataset is ready, we can save it as a CSV file in our `data`
folder.

```{r}
write_csv(surveys_complete, path = "data-processed/surveys_complete.csv")
```


## Team Challenges

These are four exercises in exploratory data analyses, which will train you to
think about data and to use the tools you have been learning about in this class
to solve scientific questions and to reproduce figures from the literature. To
solve these challenges, you will need to understand what data underlies a
figure and how you need to manipulate it to recreate the figure.


### Setting up

Start by loading the required packages. Both **`ggplot2`** and **`dplyr`** are
included in the **`tidyverse`** package collection.

```{r}
# Install if needed
# install.packages('tidyverse')
library(tidyverse)
```

### 1. Explore the weight and hindfoot trends further

Load the data we saved in the previous lesson.

```{r}
# Download if needed
# download.file("https://ndownloader.figshare.com/files/2292169", "data/portal_data.csv")
surveys <- read_csv('portal_data.csv')
surveys
```

Let's recreate the survey data set we used in lecture 4, only containing the most abundant species (those with > 800 observations).

```{r}
abundant_species <- surveys %>%
    filter(!is.na(hindfoot_length) & !is.na(weight)) %>%
    group_by(species) %>%
    tally() %>%
    arrange(desc(n)) %>%
    filter(n > 800) %>%
    select(species)

surveys_abun_species <- surveys %>%
    filter(!is.na(hindfoot_length) &
        !is.na(weight) &
        species %in% abundant_species$species)


# If everything loaded correctly, the dimensions of your data set should be
# 30,320 rows x 13 columns. We could check this by displaying the entire data
# set with `surveys_abun_species`, but also more directly by comparing the
# dimensions of the data set to those listed above.
dim(surveys_abun_species) == c(30320, 13)
# Remember that `==` is for comparisons, and `=` is for assigning arguments in
# function calls.

```

In the second question in the last challenge of lecture 4, we saw that the
average weight of all the animals decreased over time, while the average weight
for each species remained constant. Here are those plots again.

```{r}
surveys_abun_species %>%
    filter(!is.na(weight)) %>%
    group_by(year) %>%
    summarize(mean_weight = mean(weight)) %>%
    ggplot(aes(x = year, y = mean_weight)) +
        geom_line()
```

```{r}
surveys_abun_species %>%
    filter(!is.na(weight)) %>%
    group_by(year, species) %>%
    summarize(mean_weight = mean(weight)) %>%
    ggplot(aes(x = year, y = mean_weight, color = species)) +
        geom_line() +
        facet_wrap(~ species)
```

If you were to look at the average hindfoot length over time, you would find
that the trends are similar to those of the average weight. Can you find an
explanation for why both the average hindfoot length and the average weight
decrease over time for all animals' average weight, but remain constant when
looking at individual species? Phrased another way: since each species displays
constant weight and hindfoot_length measures over time, what could be the cause
of the notable decrease over time for the average weight of all species pooled
together?

```{r, eval=FALSE, echo=FALSE}
# 3.a
# Answer: Since the average hindfoot length and average weight has remained
# relatively constant for each individual species, the explaining variable is
# likely the number of species captured has shifted towards capturing more of
# the smaller rodents. By plotting the result of tallying by species, we can see
# that this is indeed the case.

surveys_abun_species %>%
    filter(!is.na(weight)) %>%
    group_by(year, species) %>%
    tally() %>%
    ggplot(aes(x = year, y = n, color = species)) +
        geom_line() +
        facet_wrap(~ species)
       
# Average weight over time
surveys_abun_species %>%
    filter(!is.na(weight)) %>%
    group_by(year, species) %>%
    summarize(mean_weight = mean(weight)) %>%
    ggplot(aes(x = year, y = mean_weight, color = species)) +
        geom_line()
       
# Total weight over time
surveys_abun_species %>%
    filter(!is.na(weight)) %>%
    group_by(year, species) %>%
    summarize(total_weight = sum(weight)) %>%
    ggplot(aes(x = year, y = total_weight, color = species)) +
        geom_line()


# 3 supplementary
# All species hindfoot length trend
surveys_abun_species %>%
    filter(!is.na(weight)) %>%
    group_by(year) %>%
    summarize(mean_hinfood_length = mean(hindfoot_length)) %>%
    ggplot(aes(x = year, y = mean_hinfood_length)) +
        geom_line()

# Per species hindfoot length trend
surveys_abun_species %>%
    filter(!is.na(weight)) %>%
    group_by(year, species) %>%
    summarize(mean_weight = mean(hindfoot_length)) %>%
    ggplot(aes(x = year, y = mean_weight, color = species)) +
        geom_line() +
        facet_wrap(~ species)
```


### 2. Reproduce figure 3 from the paper

For this section, you will apply your data wrangling and plotting skills to
reproduce a couple of figures from a study on the yearly change in biomass of
plants in the Abisko national park in northern Sweden. This paper is publish
under an open license and the [figures can be accessed via this
link](http://rstb.royalsocietypublishing.org/content/368/1624/20120486.figures-only).

You will be working with this dataset in assignment 3, so we have prepared the
data for you in a format that is easier to work with. [Download the
data](https://raw.githubusercontent.com/UofTCoders/council/master/datasets/joel/plant-biomass-preprocess.csv)
and read it into a dataframe called `plant_biomass`. Confirm that the data
frame is 180 rows by 10 columns

```{r, echo=FALSE, eval=FALSE}
plant_biomass <- read_csv('https://raw.githubusercontent.com/UofTCoders/council/master/datasets/joel/plant-biomass-preprocess.csv')
dim(plant_biomass) == c(180, 10)
```

Reproduce figure 3 from the paper. Focus on the overall message on the plot,
i.e. two panels for different habitats of the plant biomass over the year in
grazed controls vs rodent exclosures. You do not need to get the figure
aesthetics to look exactly the same as in the published figures (i.e. no need
for the exact colors and shapes, axis styles, or to include the small dots
around the main lines).

```{r, eval=FALSE, echo=FALSE}
plant_biomass %>%
    gather(species, biomass, betula_nana:vaccinium_vitis_idaea) %>%
    group_by(habitat, treatment, year) %>%
    summarize(mean_biomass = mean(biomass)) %>%
    ggplot(aes(x = year, y = mean_biomass, color = treatment)) +
        geom_line() +
        geom_point() +
        facet_grid(~ habitat)
```

### 3. Reproduce figure 4 from the paper
Use the `plant_biomass` data set and reproduce figure 4 from the paper.

Hints:

- To get the right dimensions of the subplots, explore how to use the function
  `facet_grid` instead of `facet_wrap`.
- Remember to search online for help, e.g. "How do I change the figure size in R
  markdown?"
- You will also need to search online or in the R documentation to find out how
  you change the y-scale to be constant within a species, but not between
  species (as in the paper figure).

```{r, eval=FALSE, echo=FALSE, fig.height=9, fig.width=7}
plant_biomass %>%
    gather(species, biomass, betula_nana:vaccinium_vitis_idaea) %>%
    group_by(habitat, species, treatment, year) %>%
    summarize(mean_biomass = mean(biomass)) %>%
    ggplot(aes(x = year, y = mean_biomass, color = treatment)) +
        geom_line() +
        geom_point() +
        facet_grid(species ~ habitat, scales = 'free_y')
```

### 4. Make figure 3 perfect

Let's try to bring figure 3 closer to the paper version. Use the R documentation
and search online to find out how to:

- Change the size of markers in your plots and adjust them appropriately.
- Slightly adjust the thickness of the lines.
- Change the colors of the lines to match those online.
- Apply a ggplot theme to make the figure background white and the overall
  figure appearance more like the paper version.
- Change the x and y label to match the paper figure.

```{r, eval=FALSE, echo=FALSE, fig.height=4, fig.width=10}
# There are some additional customizatios in the answer that
# I think are too much to ask for but can be nice to show
color_names <- c('red', 'blue')

plant_biomass %>%
    gather(species, biomass, betula_nana:vaccinium_vitis_idaea) %>%
    group_by(habitat, treatment, year) %>%
    summarize(mean_biomass = mean(biomass)) %>%
    ggplot(aes(x = year, y = mean_biomass, color = treatment, shape = treatment)) +
        geom_line(size = 0.8) +
        geom_point(size = 3) +
        scale_color_manual(values = color_names) +
        facet_grid(~ habitat) +
        ylab('plant biomass') +
        xlab('') +
        theme_classic() +
        theme(
            strip.background = element_blank(),
            strip.text.x = element_blank())

```


*Parts of this lesson material were taken and modified from [Data
Carpentry](http://datacarpentry.org) under their CC-BY copyright license. See
their [lesson page](http://www.datacarpentry.org/R-ecology-lesson/03-dplyr.html) for the original source.*
