
---
title: Data visualization in R
author: Lindsay Coome
---

## Lesson preamble

> ### Learning Objectives
>
> - Making better plots for outreach and communication.
> - Review how to produce scatter plots with trendlines using ggplot.
> - Make bar and other graphs using ggplot.
> - Advanced faceting and segmenting data.
> - Altering asthetic features in ggplot.
> - Making interactive and quick graphs.
> 
> ### Lesson outline
>
> - How to visualize data (20 min)
> - ggplot basics (10 min)
> - Basics and faceting (15 min)
> - Aesthetics and plot types (20 min)
> - Interactive and quick graphing (10 min)

-----

#WHAT MAKES A GOOD GRAPH?
> #### Challenge
>
> What makes a bad graph? What makes an ugly graph to you?
>


There are many considerations to take into account when creating a graph.
Data types:
* Nominal
* Ordinal
* Interval
* Ratio
And data features:
* Scalar
* Vector
* Tensor

Ask: "what kind of data/information is it?"
* Numerical?
* What type?
* Spatial?
* How many dimensions?
* Nominal/categorical?
* Narrative?

## TUFTE'S GUIDELINES
* Reduce non-data ink
* Enhance the data ink

Reduces the proportion of graphic’s ink devoted to the non-redundant display of data-information

Avoid "chartjunk" - extraneous visual elements that detract from message

Visual Attribute    | Non-Data | Data
--------------------|----------|-------
Line Thickness      | Thin     | Thick
Size                | Small    | Big
Brightness Contrast | Decrease | Increase
Colour Saturation   | Decrease | Increase

#COLOUR RULES
* Large background colors should be quiet, muted to let brighter colors stand out
* To highlight some element of a figure, using a bright colour can be effective
* However, if brightly colouring this aspect of your figure serves no purpose, leave it greyscale/plain
* Here is a reference sheet with all the colours and colour names in R:
http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf

#SUMMARY
* Determine what type of data you have and how your data will be used
* Determine how many variable types are in your data
* Decide on a visual treatment for each of the variables 
* Focus on the data in your visual representation

#MORE RESOURCES ON GENERAL PLOTTING GUIDELINES
http://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1003833
https://blog.graphiq.com/finding-the-right-color-palettes-for-data-visualizations-fcd4e707a283
http://www.cookbook-r.com/Graphs/
https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf

## WHAT IS GGPLOT?
ggplot2 is meant to be an implementation of the Grammar of Graphics, hence gg-plot. The basic notion is that there is a grammar to the composition of graphical components in statistical graphics, and by directly controlling that grammar, you can generate a large set of carefully constructed graphics tailored to your particular needs. Each component is added to the plot as a layer.

## COMPONENTS OF A GGPLOT2 PLOT
Plots convey information through various aspects of their aesthetics. Some aesthetics that plots use are:

* x position
* y position
* size of elements
* shape of elements
* color of elements

The elements in a plot are geometric shapes, like

* points
* lines
* line segments
* bars
* text

Some of these geometries have their own particular aesthetics. For instance:

**points**

* point shape
* point size

**lines**

* line type
* line weight

**bars**

* y minimum
* y maximum
* fill color
* outline color

**text**

* label value

#SETTING UP
Load the required packages:
```{r}
#library(ggplot2)
```

##LET'S GET STARTED!

Here is an example of the base plotting functions in R:
```{r cars}
plot(cars)
hist(cars$speed)
```

You've already made a few basic plots in ggplot.

We'll investigate the dataset we're going to use to make some graphs....
```{r}
library(ggplot2)
str(iris)
```

Recall:
* geom_point() for scatter plots, dot plots, etc.
* geom_line() for trend lines, time-series, etc.
* geom_histogram() for histograms

Once we've defined the structure of our graph,  we need to add a "layer" to it in order to define what type of graph it is. Let's make a scatterplot:
```{r}
ggplot(iris, aes(x=Sepal.Width, y=Sepal.Length))+
  geom_point()
```

You can also save the structure of the graph to a variable and then add layers and other attributes to it. Let's do that:
```{r}
g <- ggplot(iris, aes(x=Sepal.Width, y=Sepal.Length)) 
g + geom_point()
```

We can add other "layers" of graphs to our first graph. Let's add a trend line to our scatterplot:
```{r}
g + geom_point() + 
  geom_smooth() 
```

We can also "facet" our graphs, which means dividing a plot into subplots based on the values of one or more discrete variables.
```{r}
g + geom_point() + geom_smooth() + facet_grid(.~Species)
```

There are lots of ways to organize your facet. See "faceting" at the bottom of [ggplot cheetsheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf).

> #### Challenge
>
> Create another faceted graph using this dataset that is interpretable. Does faceting help communicate the 
> information contained in the graph? Why or why not?
>



We can play around with the labels on our facet:
```{r}
g + geom_point() + geom_smooth() + facet_grid(.~Species, labeller = label_both) +
  ylab("Sepal Length") +
  xlab ("Sepal Width")
```

Let's say we don't want to facet, but we want to colour the dots depending on the species:
```{r}
ggplot(iris, aes(x=Sepal.Width, y=Sepal.Length, color=Species)) + geom_point()+
  ylab("Sepal Length") +
  xlab ("Sepal Width")
```

We can change other aesthetic features of the graph. Let's get rid of the color aspect of our structure and change the colour and size of the points:
```{r}
ggplot(iris, aes(x=Sepal.Width, y=Sepal.Length)) + geom_point(color="steelblue", size=4)

```

Going back to our graph coloured by species, what if we want to change other aspects, like the theme, or centre the title?

```{r}
ggplot(iris, aes(x=Sepal.Width, y=Sepal.Length, color=Species)) + geom_point()+
  ylab("Sepal Length") +
  xlab ("Sepal Width") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))
  
```


What if we want a bar graph of Sepal Length by Species?
```{r}
ggplot(iris, aes(x=Species, y=Sepal.Length)) + geom_bar(stat="identity")+
  ylab("Sepal Length") +
  xlab ("Species") +
ggtitle("Sepal Width by Sepal Length") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))
```

Violin plot?
```{r}
ggplot(iris, aes(x=Species, y=Sepal.Length)) + geom_violin()+
  ylab("Sepal Length") +
  xlab ("Species") +
ggtitle("Sepal Width by Sepal Length") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))
```
