---
title: "Project and Git setup in RStudio, project work"
author: "Luke Johnston"
---

## Lesson preamble

> ### Lesson objectives:
>
> - Know why a standard/consistent folder and file structure is important
> - Learn *a bit* about how version control and Git works
> - Learn the 4 main Git commands and the Git-RStudio interface
> - Project work
>
> ### Lesson outline:
>
> - Folders and files structure (5-10 min)
>     - Using R to create this structure (5 min)
> - Version control using Git 
>     - (Basic) introduction to Git (10-15 min)
>     - Common commands for Git (15-20 min)
>     - Using Git in RStudio (10 min)
> - Quick introduction to using GitHub (5-10 min)
>     - Exercise: Fork GitHub (5-10 min)
> - Project work
>     - Get started/work on assignment 7 (20 min)

-----

## Git

### A short introduction

Include comic

What is git

Include why should use git or a version control system

Power when collaborating

A bit about how it works.

#### Exercise

Follow the getting setup steps from command line:

```bash
git config --global user.name "Your Name"
git config --global user.email yourname@domain.com
```

Replace "Your Name", with, well, your first and last name. Replace the fake
email address with the email address you normally use.

Confirm it works by running:

```bash
git config --list
```

### The basic commands of Git

- `git add`
- `git commit`
- `git log`
- `git status`

We'll switch between using the command line and using the RStudio interface.
While the RStudio interface is very useful and powerful, you will likely need to
be familiar with the command line for some Git commands.

#### Exercise

Create a new markdown file called `bio.md` in the `lec13` R project. Add your
name and one or two sentence blurb about yourself.

- *Using the command line*, add this new file and commit to the history. Check
the status and log.

- *Using the RStudio Git interface*, add the new bio file, write a short but
descriptive commit message, and hit `Commit`.

### GitHub and your project

In GitHub

```{r}
DiagrammeR::grViz("
digraph {

    node [shape = box, style = rounded]
    {
        rank = same;
        GitHubMain [label = 'Main GitHub\nrepo (UofTCoders)']
        GitHubFork [label = 'Forked GitHub\nrepo (yours)']
    }

    LocalRepo [label = 'Local repo']
    
    GitHubFork -> GitHubMain [label = 'Pull\nRequest']
    GitHubFork -> LocalRepo [label = 'remote origin']
    GitHubMain -> LocalRepo [label = 'remote upstream']
}
")
```




```{r}
DiagrammeR::grViz("
digraph {

    node [shape = box, style = rounded]
    {
        rank = same;
        GitHubMain [label = 'Main GitHub\nrepo (UofTCoders)']
        GitHubFork [label = 'Forked GitHub\nrepo (yours)']
    }

    LocalRepo [label = 'Local repo']
    
    GitHubFork -> GitHubMain [label = 'Pull\nRequest']
    GitHubFork -> LocalRepo [label = 'remote origin']
    GitHubMain -> LocalRepo [label = 'remote upstream']
}
")
```
