---
title: Data visualization in R
author: Lindsay Coome
---

## Lesson preamble

> ### Learning Objectives
>
> - Making better plots for outreach and communication.
> - Review how to produce scatter plots with trendlines using ggplot.
> - Advanced faceting and segmenting data.
> - Altering aesthetic features in ggplot.
> - Make bar and other graphs using ggplot.
> - Making interactive and quick graphs.
> - Saving graphs.
> 
> ### Lesson outline
>
> - How to visualize data (20 min)
> - ggplot basics (10 min)
> - Basics and faceting (10 min)
> - Aesthetics and plot types (25 min)
> - Interactive and quick graphing (10 min)
> - Saving graphs (5 min)
>
> ### Setup
>
> - Make sure to install `plotly` (`install.packages("plotly")`).

-----

## Making good graphs

### What makes a good graph?

#### Challenge

What makes a bad graph? What makes an ugly graph to you?

### Components of a graph: The data

There are many considerations to take into account when creating a graph.

Data types:

* Nominal (labels without quantitative values)
* Ordinal (the order of the values is what’s important, but the magnitude of
  difference between each value is not really known)
* Interval (interval scales are numeric scales in which we know not only the
  order, but also the exact differences between the values)
* Ratio (same properties as interval scales but with an absolute zero, e.g.
  height)

And data features:

* Scalar (simple magnitude or quantity)
* Vector (magnitude and direction)
* Tensor (vectors + other attributes)

Ask: "what kind of data/information is it?"

* What type?
* Numerical?
* Spatial?
* How many dimensions?
* Nominal/categorical?
* Narrative?

### Tufte's guidelines

* Reduce non-data ink
* Enhance the data ink

Reduces the proportion of graphic’s ink devoted to the non-redundant display of
data-information.

Avoid "chartjunk" - extraneous visual elements that detract from message.

Visual Attribute    | Non-Data | Data
--------------------|----------|-------
Line Thickness      | Thin     | Thick
Size                | Small    | Big
Brightness Contrast | Decrease | Increase
Colour Saturation   | Decrease | Increase

[More on Edward Tufte](https://www.edwardtufte.com/tufte/)

### Colour rules

* Large background colours should be quiet, muted to let brighter colors stand out
* To highlight some element of a figure, using a bright colour can be effective
* However, if brightly colouring this aspect of your figure serves no purpose,
  leave it greyscale/plain
* Here is a reference sheet with all the [colours and colour names in
  R](http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf)

### Summary

* Determine what type of data you have and how your data will be used
* Determine how many variable types are in your data
* Decide on a visual treatment for each of the variables 
* Focus on the data in your visual representation

### More general resources on plotting

* [Ten Simple Rules for Better Figures](http://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1003833)
* [Finding the Right Color Palettes for Data Visualizations](https://blog.graphiq.com/finding-the-right-color-palettes-for-data-visualizations-fcd4e707a283)
* [Cookbook for R - Graphs](http://www.cookbook-r.com/Graphs/)
* [ggplot cheatsheet - very useful!](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf)
* [Examples of bad graphs](https://www.biostat.wisc.edu/~kbroman/topten_worstgraphs/)
* [More examples of bad graphs and how to improve them](https://www.stat.auckland.ac.nz/~ihaka/120/Lectures/lecture03.pdf)
* [Wikipedia has a great article on misleading graphs](https://en.wikipedia.org/wiki/Misleading_graph)

## What is ggplot?

ggplot2 is inspired by the principles from the [Grammar of
Graphics](https://www.cs.uic.edu/~wilkinson/TheGrammarOfGraphics/GOG.html) and
also named thereafter. The basic notion is that there is a grammar to the
composition of graphical components in statistical graphics, and by directly
controlling that grammar, you can generate a large set of carefully constructed
graphics tailored to your particular needs. Each component is added to the plot
as a layer.

### Components of a ggplot plot

Plots convey information through various aspects of their aesthetics. Some
aesthetics that plots use are:

* x position
* y position
* size of elements
* shape of elements
* color of elements

The elements in a plot are geometric shapes, like

* points
* lines
* line segments
* bars
* text

Some of these geometries have their own particular aesthetics. For instance:

**points**

* point shape
* point size

**lines**

* line type
* line weight

**bars**

* y minimum
* y maximum
* fill color
* outline color

**text**

* label value

### Setting up

Load or install the required packages:

```{r}
library(ggplot2)
```

```{r, eval=FALSE}
install.packages("plotly")
```


### Let's get started!

Here is an example of the base plotting functions in R:

```{r cars}
plot(cars)
hist(cars$speed)
```

You've already made a few basic plots in ggplot.

We'll investigate the dataset we're going to use to make some graphs....

```{r, echo=FALSE}
iris <- datasets::iris
```

```{r}
str(iris)
```

Recall:

* geom_point() for scatter plots, dot plots, etc.
* geom_line() for trend lines, time-series, etc.
* geom_histogram() for histograms

Once we've defined the structure of our graph, we need to add a "layer" to it in
order to define what type of graph it is. Let's make a scatterplot:

```{r}
ggplot(iris, aes(x = Sepal.Width, y = Sepal.Length)) +
  geom_point()
```

You can also save the structure of the graph to a variable and then add layers
and other attributes to it. Let's do that:

```{r}
g <- ggplot(iris, aes(x = Sepal.Width, y = Sepal.Length)) 
g + geom_point()
```

We can add other "layers" of graphs to our first graph. Let's add a trend line
to our scatterplot:

```{r}
g +
    geom_point() +
    geom_smooth() 
```

We can also "facet" our graphs, which means dividing a plot into subplots based
on the values of one or more discrete variables.

```{r}
g + 
    geom_point() + 
    geom_smooth() + 
    facet_grid(. ~ Species)
```

There are lots of ways to organize your facet. See "faceting" at the bottom of
[ggplot
cheatsheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf).

Let's say we don't want to facet, but we want to colour the dots depending on
the species:

```{r}
ggplot(iris, aes(x = Sepal.Width, y = Sepal.Length, color = Species)) + 
    geom_point()
```

We can change other aesthetic features of the graph. Let's get rid of the colour
aspect of our structure and change the colour and size of the points:

```{r}
ggplot(iris, aes(x = Sepal.Width, y = Sepal.Length)) + 
    geom_point(color = "steelblue", size = 4)
```

Going back to our graph coloured by species, what if we want to change other
aspects, like the theme, or centre the title?

```{r}
ggplot(iris, aes(x = Sepal.Width, y = Sepal.Length, color = Species)) +
    geom_point() +
    ylab("Sepal Length") +
    xlab("Sepal Width") +
    ggtitle("Sepal Width by Sepal Length") +
    theme_bw() +
    theme(plot.title = element_text(hjust = 0.5))
```

What about changing the font? Let's get rid of our preconfigured theme for a
moment, which changes many aspects of the aesthetics at once. Let's spend some
time building our own theme...

```{r}
ggplot(iris, aes(x = Sepal.Width, y = Sepal.Length, color = Species)) +
    geom_point() +
    ylab("Sepal Length") +
    xlab("Sepal Width") +
    ggtitle("Sepal Width by Sepal Length") +
    theme(
        plot.title = element_text(hjust = 0.5),
        text = element_text(size = 8, family = 'Times'),
        legend.title = element_blank()
        )
```

You can add your own alterations to themes. For example, here is an APA theme
built on top of the base theme_bw(). Themes can be saved into variables so they
don't need to be typed out every time you use a graph.

```{r}
apatheme <- function() {
    theme_bw() +
        theme(
            panel.grid.major = element_blank(),
            panel.grid.minor = element_blank(),
            panel.border = element_blank(),
            axis.line = element_line(),
            text = element_text(family = 'Times'),
            legend.title = element_blank()
        )
}

ggplot(iris, aes(x = Sepal.Width, y = Sepal.Length, color = Species)) +
    geom_point() +
    ylab("Sepal Length") +
    xlab("Sepal Width") +
    ggtitle("Sepal Width by Sepal Length") +
    apatheme()
```

This is a [resource](http://ggplot2.tidyverse.org/reference/theme.html) on which
parts of the plot are controlled by different theme parameters when and if you
create your own theme.

### Bar graphs

What if we want a bar graph of Sepal Length by Species?

```{r}
ggplot(iris, aes(x = Species, y = Sepal.Length)) +
    geom_bar(stat='summary', fun.y='mean') +
    ylab("Sepal Length") +
    xlab("Species") +
    ggtitle("Sepal Length by Species") +
    theme_bw() +
    theme(plot.title = element_text(hjust = 0.5))
```

As you can see, it is a bit cumbersome to create barplots of means with ggplot
(you need to add `stat='summary', fun.y='mean'`). The default behavior of
`geom_bar` is to count observations instead of summarize them, which actually is
a more appropriate use of a barplot for reasons that we will cover below.

### Violin plots

```{r}
ggplot(iris, aes(x = Species, y = Sepal.Length)) +
    geom_violin() +
    ylab("Sepal Length") +
    xlab("Species") +
    ggtitle("Sepal Length by Species") +
    theme_bw() +
    theme(plot.title = element_text(hjust = 0.5))
```

### Box plots

We can use boxplots to visualize the distribution Sepal Length within each
species:

```{r boxplot}
ggplot(data = iris, aes(x = Species, y = Sepal.Length)) +
    geom_boxplot() +
    ylab("Sepal Length") +
    xlab("Species") +
    ggtitle("Sepal Length by Species") +
    theme_bw() +
    theme(plot.title = element_text(hjust = 0.5))
```

#### Challenge

Out of the barplot, boxplot and violinplot, which one do you think is the most
informative and why? Which is the most true to the underlying data and how
would you know this?

### Pros and cons of different graph types

We will deepen the discussion around some of these ideas, in the context of the
following plot:

![*Reproduced with permission from [Dr. Koyama's poster*](http://biostat.mc.vanderbilt.edu/wiki/pub/Main/TatsukiKoyama/Poster3.pdf)](./image/dynamite-bars.png)

Similar to the ideas of Tufte, it is generally advisable to avoid "decorative"
plot elements that do not convey extra information about the data, *especially*
when such elements hide the real data. For example, in the rectangles that make
up these bars, the only relevant information is given by the where the
rectangles ends on the y-axis, the rest of it is unnecessary and you could just
have used a marker (circle, square, etc) to indicate the height on the y-axis
instead. Note that the body of the rectangle is not representative for where the
data lies, there are probably no data points close to 0, and several above the
rectangle.

Barplots are especially misleading when used as data summaries, as in the
example above. In a summary plot, only two distribution parameters (a measure of
central tendency, e.g. the mean, and error, e.g. the standard deviation or a
confidence interval) are displayed, instead of showing all the individual data
points. This can be highly misleading, since different underlying distributions
can give rise to the same summary plot just as when we previously discussed that
the different distributions from the datasaurus R-package can give rise to [the
same linear regression coefficients](
https://uoftcoders.github.io/rcourse/lec11-modelling.html#challenge17).
We also have no idea of how many observations there are in each group. These
shortcomings become evident when comparing the barplot to the underlying
distributions that were used to create them:

![*Reproduced with permission from [Dr.
Koyama's poster*](http://biostat.mc.vanderbilt.edu/wiki/pub/Main/TatsukiKoyama/Poster3.pdf)](./image/dynamite-vs-dists.png)

Immediately, you can see that conclusions drawn from the barplot, such that A
and B have the same outcome, are factually incorrect. The distribution in D is
bimodal, so representing that with a mean would be like observing black and
white birds and conclude that the average bird color is grey, it's nonsensical.
If we would have planned our follow up experiments based on the barplot alone,
we would have been setting ourselves up for failure! Always be sceptical when
you see a barplot in a published paper, and think of how the underlying
distribution might look (note that barplots are more acceptable when used to
represents counts, proportion or percentages, where there is only one data point
per group in the data set).

Boxplots and violin plots are more meaningful data summaries as they represent
more than just two distribution parameters. However, these can still be
misleading and it is often the most appropriate to show each individual
observation with a dot/hive/swarm plot, possibly combined with with a
superimposed summary plot or a marker for the mean or median *if* you think this
conveys additional information. One exception, when it is not advisable to show
all data points, is when your data set is gigantic and plotting each individual
observation would oversaturate the chart. Here, you are forced to show summary
plots or a 2D histogram.

Here is an example of how you can plot a boxplot, together with the individual
observations:

```{r boxplot-with-points}
ggplot(data = iris, aes(x = Species, y = Sepal.Length)) +
    geom_boxplot(alpha = 0) +
    geom_jitter(alpha = 0.3, color = "tomato")
```

Notice how the boxplot layer is behind the jitter layer? What do you need to
change in the code to put the boxplot in front of the points such that it's not
hidden?

#### Challenge

So far, we've looked at the distribution of sepal length within species.  Try
making a new plot to explore the distribution of another variable within each
species.

What about making interactive graphs in R? There is a package called plotly that
lets us make our ggplot2 graphs interactive:

```{r}
library(plotly)

g <- ggplot(iris, aes(x = Species, y = Sepal.Length)) +
    geom_bar(stat='summary', fun.y='mean')
p <- ggplotly(g)
p
```

#### Challenge

Create another type of interactive graph (i.e. not a bar graph) using these
data. What kinds of graphs might interactivity add or improve the
communication and clarity of your message to?

### More resources on ggplot2

* [ggplot2 documentation](http://had.co.nz/ggplot2/)
* [Book by Hadley Wickham](https://www.amazon.com/ggplot2-Elegant-Graphics-Data-Analysis/dp/0387981403)
* [ggplot2 cheat sheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf)
* [r graph gallery for inspiration (not just limited to ggplot2 graphs)](https://www.r-graph-gallery.com/all-graphs/)

*Parts of this lesson material were taken and modified from [Software
Carpentry](https://software-carpentry.org) under their CC-BY copyright license.
See their [lesson
page](http://swcarpentry.github.io/r-novice-gapminder/08-plot-ggplot2/) for the
original source.*
